---
title: "Severe Weather Analysis"
author: "Kevin Roche"
date: "17/08/2021"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
    fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Synopsis

Using data from the [U.S. National Oceanic and Atmospheric Administration (NOAA)](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf), the following data analysis determines which weather events have been the most harmful to the US population from an economic and health perspective. 

# Setup

```{r}
## Load packages
library(tidyverse)
library(ggplot2)
```

# Data Processing

First, we need to load the data into R.

```{r}
## Load data
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
dir <- "/Users/kevinroche22/RData/SevereWeatherAnalysis"
setwd(dir)
download.file(url, "StormData.csv.bz2")
stormData <- read.csv(bzfile("StormData.csv.bz2"), header = TRUE, sep = ",")
```

Let's take a look at what we're working with.

```{r}
## Check Data
dim(stormData)
head(stormData, 3)
```
So, the data contains just over 900k observations of 37 different variables. It looks like some of the variables have missing data. Let's see how many missing values each variable has.

```{r}
## View proportions
stormData %>% summarise_each(funs(100*mean(is.na(.))))

## View totals
stormData %>% summarise_each(funs(sum(is.na(.))))
```
Interesting. The "COUNTYENDN" variable is missing entirely, while the "F" variable is ~93% missing. A very small proportion of the latitude variables are also missing. Given that I'm only interested in the variables that have economic and health consequences, these missing values won't be problematic.

Now, let's take a subset of the data that only includes the variables that are of interest to me.

```{r}
## Select Variables
stormData <- stormData %>% 
        select("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
```

Some of these variables are related. The value for "PROPDMG" doesn't represent the amount of property damage alone, it represents the coefficient - so it needs to be multiplied by the corresponding value from "PROPDMGEXP". The same logic follows for "CROPDMG" and "CROPDMGEXP". 

The code below creates two new variables, "PROPERTYDMGS" and "CROPDMGS" that multiply the values for PROPDMG and CROPDMG by their corresponding exponent. Observations that have missing exponent values or erroneous exponent values (a very small amount of the exponents are labeled "?", for example) are assigned a value of NA.

The original "PROPDMG, "PROPDMGEXP", "CROPDMG", and "CROPDMPEXP" columns are then dropped.

``` {r}
## Property Damage
stormData$PROPDMGEXP <- as.character(stormData$PROPDMGEXP)
stormData <- stormData %>% 
        mutate(PROPERTYDMGS = case_when(PROPDMGEXP == "K" ~ PROPDMG * 1000,
                                        PROPDMGEXP == "M" ~ PROPDMG * 1000000,
                                        PROPDMGEXP == "B" ~ PROPDMG * 1000000000))

## Crop Damage
stormData$CROPDMGEXP <- as.character(stormData$CROPDMGEXP)
stormData <- stormData %>% 
        mutate(CROPDMGS = case_when(CROPDMGEXP == "K" ~ CROPDMG * 1000,
                                    CROPDMGEXP == "M" ~ CROPDMG * 1000000,
                                    CROPDMGEXP == "B" ~ CROPDMG * 1000000000))

## Drop old property damage and crop damage columns
stormData <- stormData %>% 
        select(-PROPDMG, -PROPDMGEXP, -CROPDMG, -CROPDMGEXP)

## Format new columns as integers
stormData$CROPDMGS <- as.integer(stormData$CROPDMGS)
stormData$PROPERTYDMGS <- as.integer(stormData$PROPERTYDMGS)
```

The date variable, BGN_DATE, isn't formatted as a date. Let's fix that, and then extract the year from it. 

```{r}
## Format as date
stormData$BGN_DATE <- as.Date(stormData$BGN_DATE, "%m/%d/%Y")

## Create new variable, year
stormData <- stormData %>% 
        mutate(YEAR = str_extract(BGN_DATE, "^\\d{4}")) %>%
        select(-BGN_DATE)

## Format year variable
stormData$YEAR <- lubridate::ymd(stormData$YEAR, truncated = 2L)
```

The NOAA [notes](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) that prior to 1996, only four types of events were recorded - Hail, Wind, Thunderstorms, and Tornadoes. From 1996 onward, 48 event types are recorded. The data will be skewed towards the original four events if I use data from all time periods, so I'll need to subset the data from 1996 onward to allow for proper comparison between event types.

```{r}
## Filter on observations from 1996 onward
stormData <- stormData %>% 
        filter(YEAR >= "1996-01-01")
```

We only need events that resulted in either health or economic impacts, so we can filter out observations that don't fall into either of those categories.

```{r}
## Only keep events with health or economic impacts
stormData <- stormData %>% 
        filter(PROPERTYDMGS > 0 | CROPDMGS > 0 | FATALITIES > 0 | INJURIES > 0)
```

Let's check the cleanliness of the event type data.

```{r}
## Change case of all event types to upper
stormData$EVTYPE <- toupper(stormData$EVTYPE)
stormData %>% distinct(EVTYPE)
```
Oof. There are 186 observations, when the NOAA says there should only be 48. This means there are a lot of duplicates in there.

Because I'm looking for the most harmful events - and, I suppose, to avoid large amounts of tedious work - I'm going to only clean observations that are in the 90th percentile as far as health and economic impacts go.

```{r}
## Change NA's to 0's
stormData <- stormData %>% 
        mutate(CROPDMGS = replace_na(CROPDMGS, 0))
stormData <- stormData %>% 
        mutate(PROPERTYDMGS = replace_na(PROPERTYDMGS, 0))

## Create variables representing health and economic impacts
stormData <- stormData %>% 
        mutate(HEALTHIMPACT = INJURIES + FATALITIES, ECONIMPACT = PROPERTYDMGS + CROPDMGS)

## View observations in 90th percentile of health impacts
stormData %>% 
        group_by(EVTYPE) %>% 
        summarise(HEALTHIMPACT = sum(HEALTHIMPACT)) %>% 
        arrange(desc(HEALTHIMPACT)) %>% 
        filter(HEALTHIMPACT > quantile(HEALTHIMPACT, 0.9))

## View observations in 90th percentile of economic impacts
stormData %>% 
        group_by(EVTYPE) %>% 
        summarise(ECONIMPACT = sum(ECONIMPACT)) %>% 
        arrange(desc(ECONIMPACT)) %>% 
        filter(ECONIMPACT > quantile(ECONIMPACT, 0.9))

## Clean 90th percentile observations
stormData <- stormData %>% 
        mutate_at("EVTYPE", str_replace, "TSTM WIND", "THUNDERSTORM WIND") %>% 
        mutate_at("EVTYPE", str_replace, "RIP CURRENTS", "RIP CURRENT") %>% 
        mutate_at("EVTYPE", str_replace, "FOG", "FREEZING FOG") %>% 
        mutate_at("EVTYPE", str_replace, "WILD/FOREST FIRE", "WILDFIRE") %>%
        mutate_at("EVTYPE", str_replace, "HURRICANE/TYPHOON", "HURRICANE (TYPHOON)") %>% 
        mutate_at("EVTYPE", str_replace, "EXTREME COLD/WIND CHILL", "EXTREME COLD") %>% 
        mutate_at("EVTYPE", str_replace, "HURRICANE$", "HURRICANE (TYPHOON)")
```
At this point, the data set is clean enough to conduct some analysis on.

## Results

Let's visualize which weather events had the biggest health impacts.

```{r}
## Events with the largest total health costs
healthCosts <- stormData %>% 
        group_by(EVTYPE) %>% 
        summarise(HEALTHIMPACT = sum(HEALTHIMPACT)) %>% 
        arrange(desc(HEALTHIMPACT)) %>%
        top_n(10) %>% 
        ggplot(aes(x = reorder(EVTYPE, HEALTHIMPACT), y = HEALTHIMPACT, color = EVTYPE)) +
        geom_bar(stat="identity") + 
        theme_bw() +
        theme(legend.position="none") +
        ggtitle("Health Costs (Injuries + Fatalities) by Event Type") +
        labs(x = "Event Type", y = "Total Health Impacts") +
        coord_flip()
        
## Events with the largest number of injuries
injuries <- stormData %>% 
        group_by(EVTYPE) %>% 
        summarise(INJURIES = sum(INJURIES)) %>% 
        arrange(desc(INJURIES)) %>%
        top_n(10) %>% 
        ggplot(aes(x = reorder(EVTYPE, INJURIES), y = INJURIES, color = EVTYPE)) +
        geom_bar(stat="identity") + 
        theme_bw() +
        theme(legend.position="none") +
        ggtitle("Injuries by Event Type") +
        labs(x = "Event Type", y = "Total Injuries") +
        coord_flip()

## Events with the largest number of deaths
deaths <- stormData %>% 
        group_by(EVTYPE) %>% 
        summarise(FATALITIES = sum(FATALITIES)) %>% 
        arrange(desc(FATALITIES)) %>%
        top_n(10) %>% 
        ggplot(aes(x = reorder(EVTYPE, FATALITIES), y = FATALITIES, color = EVTYPE)) +
        geom_bar(stat="identity") + 
        theme_bw() +
        theme(legend.position="none") +
        ggtitle("Fatalities by Event Type") +
        labs(x = "Event Type", y = "Total Deaths") +
        coord_flip()

## Plot
gridExtra::grid.arrange(healthCosts, injuries, deaths, heights = unit(c(2.5,2.5,2.5), c("in", "in", "in")), nrow = 3)
```

We observe that tornadoes lead to the most injuries, while excessive heat causes the most fatalities. 

Now, let's visualize which events had the biggest economic consequences.

```{r}
## Events with the largest total economic costs
econData <- stormData %>% 
        group_by(EVTYPE) %>% 
        summarise(ECONIMPACT = sum(ECONIMPACT)) %>% 
        arrange(desc(ECONIMPACT)) %>% 
        top_n(10) %>% 
        ggplot(aes(x = reorder(EVTYPE, ECONIMPACT), y = ECONIMPACT, color = EVTYPE)) +
        geom_bar(stat="identity") + 
        theme_bw() +
        theme(legend.position="none") +
        ggtitle("Economic Costs (Property + Crop Damage) by Event Type") +
        labs(x = "Event Type", y = "Total Economic Costs") + 
        coord_flip() +
        scale_y_continuous(labels=scales::dollar_format())

## Events with the largest total property damage
propertyDamage <- stormData %>% 
        group_by(EVTYPE) %>% 
        summarise(PROPERTYDMGS = sum(PROPERTYDMGS)) %>% 
        arrange(desc(PROPERTYDMGS)) %>% 
        top_n(10) %>% 
        ggplot(aes(x = reorder(EVTYPE, PROPERTYDMGS), y = PROPERTYDMGS, color = EVTYPE)) +
        geom_bar(stat="identity") + 
        theme_bw() +
        theme(legend.position="none") +
        ggtitle("Property Damage by Event Type") +
        labs(x = "Event Type", y = "Total Property Damage") + 
        coord_flip() +
        scale_y_continuous(labels=scales::dollar_format())

cropDamage <- stormData %>% 
        group_by(EVTYPE) %>% 
        summarise(CROPDMGS = sum(CROPDMGS)) %>% 
        arrange(desc(CROPDMGS)) %>% 
        top_n(10) %>% 
        ggplot(aes(x = reorder(EVTYPE, CROPDMGS), y = CROPDMGS, color = EVTYPE)) +
        geom_bar(stat="identity") + 
        theme_bw() +
        theme(legend.position="none") +
        ggtitle("Crop Damage by Event Type") +
        labs(x = "Event Type", y = "Total Crop Damage") + 
        coord_flip() +
        scale_y_continuous(labels=scales::dollar_format())

## Plot
gridExtra::grid.arrange(econData, propertyDamage, cropDamage, heights = unit(c(2.5,2.5,2.5), c("in", "in", "in")), nrow = 3)
```

We observe that floods cause the most property damage, while droughts cause the most crop damage. This makes intuitive sense - floods are somewhat common in homes, and crops often die in a drought.